Querying Integrated Scientific Observation and Measurement Data

### 2010-04-29 docs/algdraft/DataIntegration/DataIntegration.tex
Refine the model and query rewriting.

\title{Querying Integrated Ecological Data}

@@

In the ecology domain, most data people collected are observation and
measurement data. {\bf Add one sentence to explain this.}
Naturally, when searching ecological data, people are more interested in finding datasets
related to specific observation and measurement. 
In this section, we first illustrate the data model. Then, we
formalize the queries that scientists in this domain tend to ask.

\subsection{Data model}

@@

\subsection{Annotation}
In the Ecology domain, people are using annotations to connect
datasets to the OBOE model. 
{\bf TODO: Put a screen-dump of annotation.}

@@

{\bf TODO:} Describe the constraints in annotation

\subsection{Observation and measurement query}

{\bf TODO: Have a running example from.}
Observation and measurement (OM) query

@@

For query rewriting, the first step is to find the candidate data
tables and the attribute name lists that can be used to answer our queries. 
To find the candidate data table, the metadata tables and annotation
tables are searched. 
When the candidate data tables and the needed attributes are known, 
the second step translate the needed queries over the candidate data
tables.

When a query {\tt concept:cond} is given, the {\tt concept} can only be interpreted using
annotations. When seeing a basic concept (observation, measurement,
characteristic, protocal), their related concepts information can be
retrieved through $A.MeasType$, $A.ObsType$.

From the {\em measuremen type} and $A.Map$, we can get the attribute
names; Then, using the attribute names and the metadata table (with
table and column definition), we can get the candidate data tables
with their corresponding attribute lists.

If there is context, we need to use $A.ContextType$.

If there is distinct constraint, need to find the key measurements. 
then, perform ``group by'' operation.

@@

The information that users can have: metadata about the table.

In the scientific data, it is unrealistic for the users to know the
table structure to perform the query.

Use the materialized database, we would have four tables for all the datasets: 
entity, observation, measurement, context. 
and 4 tables for entity types, observation types, measurement types,
and context relationships.

To answer KQ1: it's simple join between

\begin{example}
Take KQ1 as example. 
concept = observation, cond is {\em species = ``Picea rubens''};

>>>>

\title{Querying Integrated Scientific Observation and Measurement Data}

@@

In many scientific domains (e.g., ecology, hydrology, earth science, geology), people collect observation and
measurement data. Such data record the observed value of some real
world entity at some specific place. E.g., ecologist studying the
growth pattern and treatments often need to record the tree height,
The collected data reflect the fact the tree height of a specific tree is 1.0in on
May 1, 2009 and 1.5in on May 1, 2010. ({\bf check whether this is
  realistic or not.})

In scientific domains where people collect observation and measurement data, there are
several commonly used and widely recognized canonical
concepts(\cite{oboe, om}). 
In this paper, we refer these concepts as OM concetps. 
The canonical concepts include {\em observation}, {\em measurement},
{\em characteristic}, {\em standard}, {\em protoca} or ({\em procedure}). 
For example, {\bf add an example to illustrate these concepts. }

When a user poses a query, it is unrealistic for a him/her to know the 
underlying data structure of the data.  
But they all know the well recognized OM conceps as discussed above. 
So, naturally, when searching such scientific data, people are more interested in finding datasets
related to such concepts, e.g., specific observation and measurement.

In this section, we first illustrate the data model. Then, we
formalize the queries that scientists in this domain tend to ask.

\subsection{Data model}\label{sec:datamodel}

@@

\subsection{Annotation}\label{sec:annotation}.
In the Ecology domain, people are using annotations to connect
datasets to the OBOE model. 
{\bf TODO: Put a screen-dump of annotation.}
When some annotations are done over data tables, the system keeps the
information $\langle AnnotId, DTableId\rangle$.

@@

{\bf TODO: Describe the constraints in annotation}

\subsection{Observation and measurement (OM) query}

{\bf TODO: Have a running example from here.}
Observation and measurement (OM) query

@@

As described in Section \ref{sec:dataquery}, 
a user's query generally does not contain the underlying data
structure because the user cannot get such information. 
On the contrary, they formulate their query with the well recognized
OM concept terminologies such as observation
measurement, characteristic, etc. 
To answer such query from the original data model (i.e., dataset
metadata and data tables), we need to {\em rewrite} the quey over the
real data table. 
In what follows, we first sketch the re-writing process. Then, we
detailed the procedure in using the data model and
structures. Finally, we include the process for the more complicated
cases (e.g., with distinct, aggregate).

Roughly speaking, the query rewriting consists of two steps:
\begin{itemize}
\item From the given query, find the relevant data tables and
  attributes that need to be used answer the given query. 
\item Translate the given query to queries over the relevant data
  tables. 
\end{itemize}

{\bf TODO: put a figure here to illustrate the steps}.

The first step is to map the OM query to the real data structure. 
We can utilize the annotation structure $A.Map$ which keeps the correspondences between
measurement type and table attributes. 
So, when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $AnnotId$. 
In case that the given OM concept is a non-measurement concept, we can use
other Annotation structures $A.*$ to figure out the measurementt types. 
E.g., if the OM concept is Characteristic ``Species'', we can use
$A.MeasType$ to get the measurement type id.

As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
information $\langle AnnotId, DTableId\rangle$. 
With the $AnnotId$, we can ge the data table information from the
metadata  $\langle AnnotId, DTableId\rangle$. 
Then, we get the relevant data table information and the needed attribute
information.

Let $RD$ be the set of relevant data tables and $d_{a_1} \cdots d_{a_m}$
are the related attributes for table $d\in RD$. We can translate the original OM query to SQL queries
over the data tables.

{\tt 
SELECT DISTINCT d.record\_id\\
FROM d\\
WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
$[$GROUP BY$]$\\
$[$HAVING$]$
}

The above describes the process for the most basic OM query. 
However, in the real application, people ask queries with distinct
observation or entity constraints. Also, people ask queries about
different observatitons using context.

{\bf TODO: Discuss how to deal with the distinct constraints. }
If there is distinct constraint, need to find the key measurements. 
then, perform ``group by'' operation.

{\bf TODO: Discuss how to deal with context}. 
If there is context, we need to use $A.ContextType$.

% Use the materialized database, we would have four tables for all the datasets: 
% entity, observation, measurement, context. 
% and 4 tables for entity types, observation types, measurement types,
% and context relationships.

\begin{example}
{\bf TODO: refine this example}
Take KQ1 as example. 
concept = observation, cond is {\em species = ``Picea rubens''};

@@
